A global compilation reveals divergent biomass partitioning between plant functional types

### 2014-10-29 manuscript/draftpaper.Rnw
one ref
Major update mixmean, incorporates significance letters, and meansbypft (slightly more generic and clean). Major update FigS8 (all pft-biome combos included when data present; caption update).

Major plant functional types differ widely in leaf mass per area (SLM, the ratio of leaf mass to area, an index of leaf thickness) (\cite{poorter_causes_2009}), which can be expected to affect biomass partitioning in several ways. First, SLM correlates with leaf lifespan. Longer lived leaves would lead to higher plant leaf mass ($M_F$) as long as annual production rates are similar (\cite{enquist_global_2002}). Another interpretation is that plants may target a certain leaf area display given a certain biomass investment in supporting tissues, because light interception and photosynthesis correlate with the exchange area of the plant.

It has been reported numerous times that gymnosperms (with higher SLM) have higher leaf mass fraction than angiosperms (with lower SLM) \cite{enquist_global_2002, cheng_relationships_2007, mccarthy_consistency_2007, poorter_biomass_2012, sardans_tree_2013}, although a recent study reported no difference (\cite{wolf_allometric_2010}). No study to date has been able to further tease apart PFT differences, or test the potentially confounding effects of climate.

@@

We then expressed the average $M_F/A_S$ as a function of the leaf mass per area (SLM = $M_F / A_F$), a measure of the leaf construction cost to the plant. We found that, across PFTs, $M_F/A_S$ was approximately proportional to the average leaf mass per area for the PFT. As a result of this approximate proportionality, the ratio of leaf area to basal stem area ($A_F/A_S$) was not different between PFTs (Fig.~\ref{fig:figure4}) (see also Fig.~\ref{fig:figureSI6}). This relationship also held approximately when data were first averaged by species (Fig.~\ref{fig:figureSI7}).

Because there was substantial variation between individual plants (Figs.~\ref{fig:figure3}), we estimated the variance explained by PFT, plant height, and biome in $M_F/A_S$ using a linear mixed-effects model (see Table~\ref{varpartpipe}). We found that plant height and biome explained only \Sexpr{100*signif(Table_pipemodel_varpart$Marginal[2],2)}\%, but when PFT was added to the model, the variance explained increased to \Sexpr{100*signif(Table_pipemodel_varpart$Marginal[3],2)}\%. If we used SLM as a predictor instead of PFT, the $R^2$ was similar (\Sexpr{100*signif(Table_pipemodel_varpart$Marginal[4],2)}\%), showing that PFT affected $M_F/A_S$ largely through differences in SLM. Consistent with this, PFT, plant height and biome explained very little variance in $A_F/A_S$ (Table~\ref{varpartpipe}).

We also partionined the variance of LMF and LAR (Table~\ref{varpartpipe}). When added to a model that included plant height and biome, PFT increased the variance explained in log-transformed LMF from \Sexpr{100*signif(Table_LMFLAR_varpart$Marginal[2],2)} to \Sexpr{100*signif(Table_LMFLAR_varpart$Marginal[3],2)}\%. However, SLM was not a good substitute for PFT, as the explained variance decreased to \Sexpr{100*signif(Table_LMFLAR_varpart$Marginal[4],2)}\%. For (log-transformed) LAR, PFT and SLM explained very little additional variance.

@@

@@

How is allocation implemented in current DGVMs? Either constant allocation fraction, target allometric equations, or optimization \cite{de2014does}. Constant allocation fraction gives constant LAR by pft if leaf lifespan is exactly proportional to SLM. Target allometric equations can be made pft-invariant by using allometry of leaf area, not leaf mass. Optimization models probably already give the right answer because normally LAI is optimized (Woodward).

@@

For the aboveground biomass pools, we calculated the leaf mass fraction ($M_F / M_T$, where $M_T$ is total aboveground biomass) and leaf area ratio ($A_F / M_T$). LMF, LAR, SLM, $M_F$ and $A_F$ are related by,

$$LMF = M_F / M_T = (M_F / A_F) (A_F / M_T) = SLM * LAR$$

We were unable to further decompose $A_S/M_W$ into wood volume and wood density components (cf. \cite{chave_tree_2005}) because wood density data were available only for a small subset of data.
We did not include SLM estimates from another database (such as TRY), instead only used SLA directly estimated for the harvested plants (typically for a subsample of leaves). For conifers, leaf area was converted to half-total surface area, which is most appropriate in terms of light interception  \cite{lang_application_1991, chen_defining_1992}, for comparison to flat leaves.

@@

<<results='asis'>>=
load("tables/Tables_varpart.RData")
print(xtable(Table_LMFLAR_varpart, label="varpartlmflar",
      caption="Explained variance in the leaf mass fraction (leaf mass / aboveground biomass) and leaf area ratio (leaf area / aboveground biomass). See Table~\\ref{varpartpipe} for more explanation. For plant height, we also added the squared term as a fixed effect, and the interaction with PFT or SLM, because the relationship was non-linear."),
      include.rownames=FALSE, caption.placement="top")
@

@@

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.99\textwidth]{FigureSI-8_mlf_alf_astbaest_pftlongmeans.pdf}
    \caption{Like Fig.~\ref{fig:figure2}, but for five combinations of major plant functional type and biome. Out of 9 possible biome / pft combinations, only five in the dataset had leaf area estimates (and two combo's are not in the dataset, e.g. tropical evergreen gymno). Note that within pft, vegetation is never significant (this result will also be in another figure!).}
    \label{fig:figureSI8}
\end{figure}

>>>>

Major plant functional types differ widely in leaf mass per area (LMA, the ratio of leaf mass to area, an index of leaf thickness) (\cite{poorter_causes_2009}), which can be expected to affect biomass partitioning in several ways. First, LMA correlates with leaf lifespan. Longer lived leaves would lead to higher plant leaf mass ($M_F$) as long as annual production rates are similar (\cite{enquist_global_2002}). Another interpretation is that plants may target a certain leaf area display given a certain biomass investment in supporting tissues, because light interception and photosynthesis correlate with the exchange area of the plant.

It has been reported numerous times that gymnosperms (with higher LMA) have higher leaf mass fraction than angiosperms (with lower LMA) \cite{enquist_global_2002, cheng_relationships_2007, mccarthy_consistency_2007, poorter_biomass_2012, sardans_tree_2013}, although a recent study reported no difference (\cite{wolf_allometric_2010}). No study to date has been able to further tease apart PFT differences, or test the potentially confounding effects of climate.

@@

We then expressed the average $M_F/A_S$ as a function of the leaf mass per area (LMA = $M_F / A_F$), a measure of the leaf construction cost to the plant. We found that, across PFTs, $M_F/A_S$ was approximately proportional to the average leaf mass per area for the PFT. As a result of this approximate proportionality, the ratio of leaf area to basal stem area ($A_F/A_S$) was not different between PFTs (Fig.~\ref{fig:figure4}) (see also Fig.~\ref{fig:figureSI6}). This relationship also held approximately when data were first averaged by species (Fig.~\ref{fig:figureSI7}).

Because there was substantial variation between individual plants (Figs.~\ref{fig:figure3}), we estimated the variance explained by PFT, plant height, and biome in $M_F/A_S$ using a linear mixed-effects model (see Table~\ref{varpartpipe}). We found that plant height and biome explained only \Sexpr{100*signif(Table_pipemodel_varpart$Marginal[2],2)}\%, but when PFT was added to the model, the variance explained increased to \Sexpr{100*signif(Table_pipemodel_varpart$Marginal[3],2)}\%. If we used LMA as a predictor instead of PFT, the $R^2$ was similar (\Sexpr{100*signif(Table_pipemodel_varpart$Marginal[4],2)}\%), showing that PFT affected $M_F/A_S$ largely through differences in LMA. Consistent with this, PFT, plant height and biome explained very little variance in $A_F/A_S$ (Table~\ref{varpartpipe}).

We also partionined the variance of LMF and LAR (Table~\ref{varpartpipe}). When added to a model that included plant height and biome, PFT increased the variance explained in log-transformed LMF from \Sexpr{100*signif(Table_LMFLAR_varpart$Marginal[2],2)} to \Sexpr{100*signif(Table_LMFLAR_varpart$Marginal[3],2)}\%. However, LMA was not a good substitute for PFT, as the explained variance decreased to \Sexpr{100*signif(Table_LMFLAR_varpart$Marginal[4],2)}\%. For (log-transformed) LAR, PFT and LMA explained very little additional variance.

@@

More plasticity in LMA than biomass partitioning has been demonstrated \cite{poorter_biomass_2012}.

@@

How is allocation implemented in current DGVMs? Either constant allocation fraction, target allometric equations, or optimization \cite{de2014does}. Constant allocation fraction gives constant LAR by pft if leaf lifespan is exactly proportional to LMA. Target allometric equations can be made pft-invariant by using allometry of leaf area, not leaf mass. Optimization models probably already give the right answer because normally LAI is optimized (Woodward).

@@

For the aboveground biomass pools, we calculated the leaf mass fraction ($M_F / M_T$, where $M_T$ is total aboveground biomass) and leaf area ratio ($A_F / M_T$). LMF, LAR, LMA, $M_F$ and $A_F$ are related by,

$$LMF = M_F / M_T = (M_F / A_F) (A_F / M_T) = LMA * LAR$$

We were unable to further decompose $A_S/M_W$ into wood volume and wood density components (cf. \cite{chave_tree_2005}) because wood density data were available only for a small subset of data.
We did not include LMA estimates from another database (such as TRY), instead only used SLA directly estimated for the harvested plants (typically for a subsample of leaves). For conifers, leaf area was converted to half-total surface area, which is most appropriate in terms of light interception  \cite{lang_application_1991, chen_defining_1992}, for comparison to flat leaves.

@@

<<results='asis'>>=
load("tables/Tables_varpart.RData")
print(xtable(Table_LMFLAR_varpart, label="varpartlmflar",
      caption="Explained variance in the leaf mass fraction (leaf mass / aboveground biomass) and leaf area ratio (leaf area / aboveground biomass). See Table~\\ref{varpartpipe} for more explanation. For plant height, we also added the squared term as a fixed effect, and the interaction with PFT or LMA, because the relationship was non-linear."),
      include.rownames=FALSE, caption.placement="top")
@

@@

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.99\textwidth]{FigureSI-8_mlf_alf_astbaest_pftlongmeans.pdf}
    \caption{Mean leaf mass per unit basal stem area ($M_F/A_S$) and leaf area per unit basal stem area ($A_F/A_S$) by PFT and major biome. No leaf area estimates were available for boreal deciduous angiosperms. DA = deciduous angiosperm, EA = evergreen angiosperm, EG = evergreen gymnosperm, Trop. = tropical, Temp. = temperate. Error bars are 95\% confidence intervals for the mean, estimated with a linear mixed-effects model using species within study as the random effect. Letters denote significant differences (at $\alpha = 0.05$)}
    \label{fig:figureSI8}
\end{figure}
